---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tAUCMNAR

<!-- badges: start -->
<!-- badges: end -->

This package addresses the challenge of handling nonignorable missingness in biomarker values when
evaluating potential heterogeneity in the discriminatory performance of biomarkers, based on the manuscript "Addressing Nonignorable Missing Data and Heterogeneity in Prognostic Biomarker Assessment".


## Installation

You can install the development version of tAUCMNAR from [GitHub](https://github.com/) with:


``` {r eval = F}
install.packages("devtools")
devtools::install_github("xinran-h/tAUCMNAR")
```

## tAUCMNAR in a nutshell
This package has four functions, est.pi.func, data_crossingdc, covariance_cal, and auc_pred.

### Function: est.pi.func
This function is used to perform two-step GMM estimation for the missing data logistic regression model. We consider the case where there is only one instrumental variable, which is either continuous or discrete. The arguments of this function are listed below:

-  `x`  A numeric matrix, including the design matrix in the missing data logistic regression model.
-  `z_cont` A vector of column names from `x` representing continuous covariates that are non-instrumental variables. If there is no continuous covariate, set it as NULL.
-  `z_dis` A vector of column names from `x` representing categorical covariates that are non-instrumental variables. If there is no categorical covariate, set it as NULL.
-  `u_type` A character string indicating the type of instrumental variable, set to "continuous" or "dicrete".
-  `u` A numeric vector representing the instrumental variable. 
-  `V` A numeric vector of missing data indicators, `V` = 1 if observed and `V` = 0 if missing.
-  `futime` A numeric vector of observed event time.
-  `status` A numeric vector of censoring indicator, status = 1 if event occurs and status = 0 if censored.
-  `q` a numeric value indicating the number of estimating equations.
-  `initial` A numeric vector of initial values for the parameters in the missing data logistic regression model. This is used in the optim() function to be optimized over.
-  `method` A character string indicating the optimization method, set to  one of the "Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN","Brent".

This function returns a list of parameter estimate, variance covariance matrix, as well as intermediate results which will be used later to obtain the asymptotic variance of the paramters in the AUC regression model. Specifically, the list contains the following items:

- `phi.hat` Estimated regression coefficients from the missing data model.
- `convergence` Convergence code from the second step optimization, see optim() for details.
- `convergence.message` Convergence message from the second step optimization, see optim() for details.
- `objective`  Value of the objective function from the second step optimization to be minimized over
- `vcov.hat` Variance-covariance matrix for the parameter estimates in the missing data logistic regression model. 
- `pi.hat` Estimated probability of observing the biomarker.
- `h.hat` A numeric matrix with n rows and q columns, each row is \eqn{h(\hat{\phi},y_i)}. 
- `composit` This is \eqn{{(\Gamma^T{\Omega}^{-1}\Gamma)}^{-1}\Gamma^T{\Omega}^{-1}}.


### Function: data_crossingdc
The function data_crossingdc is used to reshape data and leverage the existing software for fitting generalized linear model (GLM) for estimation. See Algorithm 1 in the follwoing paper:Jiang X, Li W, Li R, Ning J. Addressing subject heterogeneity in time-dependent discrimination for biomarker evaluation. Stat Med. 2024;43(7):1341-1353. doi:10.1002/sim.10024. The arguments of this function are listed below:

- `V`   A numeric vector of missing data indicators, V = 1 if observed and V = 0 if missing
- `vt`  A numeric vector  of observed time
- `vc` A numeric vector of censoring indicator
- `vm` A numeric vector of biomarker values
- `vxs` A covariate matrix used for the AUC logistic regression, do not include intercept or observed time
- `pi` A numeric vector containing the estimated probability of observing biomarker values
- `cont_vars` A vector of column names from vxs representing continuous covariates
- `discrete_vars` A vector of column names from vxs representing discrete covariates
- `c0` A constant value used for the bandwidth h, h = c0*n^(-1/3)

This function returns a dataframe where each row corresponds to a case-control pair at each observed event time. The dataframe has the following components:

- `i` A numeric indices for cases
- `idi` Patient numeric id for cases
- `idl` Patient numeric id for controls in the risk set
- `yi` Observed event time for cases
- `yl` Observed event time for controls in the risk set
- `Iil` Indicator for comparison of baseline biomarker values (1 if case biomarker > control biomarker, 0 otherwise)
- `pii` Estimated probability of observing the biomarker for cases
- `pil` Estimated probability of observing the biomarker for controls in the risk set
- `event_indicator_i` Censoring indicator for cases, this is always 1
- `event_indicator_l` Censoring indicator for controls in the risk set, 1 if event occurred, 0 if censored
- Covariates with "i" suffix Covariate columns, with names based on the original covariate name and an "i" suffix for case values
- `rho01` and `rho02` Numeric vectors representing concordant and discordant event values, respectively, when continuous covariates are present
- `rhoweight` Kernel weight vector, calculated when continuous covariates are present


### Function: covariance_cal_wrapper
The function `covariance_cal_wrapper` is a wrapper function that calls the c++ function covariance_cal, which computes three matrices: \(\Sigma_1\), \(\Sigma_2\), and \(V\). The last item is the estimated asymptotic variance for AUC logistic regression model parameters. The arguments of this function are listed below:

- `a` A numeric vector of regression coefficients from the AUC logistic regression model
- `b` A numeric vector of biomarker values
- `c` The `rho01` vector from the dataframe returned by the `data_crossingdc` function. When there are no continuous variables, use the `Iil` vector from the dataframe returned by the `data_crossingdc` function
- `d` The `rho02` vector from the dataframe returned by the `data_crossingdc` function. When there are no continuous variables, use the `1 - Iil` vector from the dataframe returned by the `data_crossingdc` function
- `e` A numeric matrix with k rows and n columns. Each row represents a predictor used in the AUC logistic regression model, including the intercept as the first column. The number of columns is equal to the number of rows in the dataframe returned by the `data_crossingdc` function 
- `f`: The `idi - 1` vector from the dataframe returned by the `data_crossingdc` function
- `g`: The `idl - 1` vector from the dataframe returned by the `data_crossingdc` function
- `h` A numeric matrix with q rows and n columns. Each row represents a predictor used in the missing data logistic regression model for cases, including the intercept as the first column. The number of columns is equal to the number of rows in the dataframe returned by the `data_crossingdc` function 
- `i` A numeric matrix with q rows and n columns. Each row represents a predictor used in the missing data logistic regression model for controls in the risk set, including the intercept as the first column. The number of columns is equal to the number of rows in the dataframe returned by the `data_crossingdc` function 
- `j` A vector of estimated probability of observing biomarkers among case patients
- `k` A vector of estimated probability of observing biomarkers among control patients in the risk set
- `l` A vector of regression coefficients from the missing data model
- `m` A numeric matrix, returned from the `est.pi.func` function, accessed through the `composit` element
- `n` A numeric matrix,, returned from the `est.pi.func` function, accessed through the `h.hat` element.


This function returns a list containing:

- `sigma1` The matrix \(\Sigma_1\), used in calculating the asymptotic variance
- `sigma2` The matrix \(\Sigma_2\), also used in calculating the asymptotic variance
- `V` The asymptotic variance \(V\) for the regression coefficients


### Function: auc_pred
The function auc_pred is to estimate the the time-dependent AUC with 95% confidence interval. The arguments of this function are listed below:

- `beta.hat`   Estimated regression coefficients from the AUC regression model
- `V`  The estimated asymptotic variance from the AUC regression model
- `tt` A vector of time
- `coeffs` A vector of covariate values, do not include intercept or observed time, corresponds to the order in the `beta.hat` 
- `nf` A numeric value of 3 or 7 for the number of polynomials. When `nf = 7`, the vector of polynomials is $\text{c}(t^{-2}, t^{-1}, t^{-0.5}, \log(t), t^{0.5}, t, t^2)$. When `nf = 3`, the vector of polynomials is $\text{c}(t^{0.5}, t, t^2)$

This function returns a matrix of 95% confidence intervals for the time-dependent AUC, where the first row represents point estimate, the second row is the lower level, and the third row is the upper level.




## An example

We use a demo data to illustrate how to use this package. The demo data can be loaded by the following code.
```{r, eval = F}
library(tAUCMNAR)
dd = tAUCMNAR::demo
```

This data is a dataframe of 1000 rows and 7 columns:`ID`, `Y`, `delta`, `z`, `u`,`M`, and 'V'. The columns are defined as follows:

- `ID` Patient ID
- `Y` Observed event time
- `delta` Censoring indicator, 0 = censored, 1 = event 1, 2 = event 2
- `z` A numeric vector indicating a continuous non-instrument covariate
- `u` A numeric vector indicating a discrete instrument covariate
- `M` Baseline biomarker values
- `V` A missing indicator for biomarker values, V = 1 if observed, V = 0 if missing

The biomarker `M` is missing for some patients. We model the probability of the biomarker being
observed using the following logistic regression:

\[ logit\{\pi_i(\boldsymbol{\phi})\} = \phi_0 + \phi_1 z_{i} + \phi_2 M_i +  \phi_3 Y_i + \phi_4 \delta_i \].
We estimate the coefficients using two-step GMM estimation based on the unbiased estimating equations:
\[  h_n(\boldsymbol{\phi}) = \sum_{i=1}^n \left [ \boldsymbol{\eta}_i \left \{V_i/\pi_i(\boldsymbol{\phi}) - 1\right \}\right]/n, \]
where \(\boldsymbol{\eta}_i = (z_{i}, u_i, 1-u_i, Y_i, \delta_i)\). The following code estimates the probability of observing the biomarker for each patient and stores the result in a list named `phi.info`.
```{r, eval  = F}

# construct a numeric matrix containing the design matrix in the logistic regression model 
 n = nrow(dd)
 x <- data.matrix(cbind(int=rep(1,n), dd[, !names(dd) %in% c("id","V","u")] ))
 
 # reorder
 x <- x[,c(1,4,5,2,3)]    
 
# perform two-step GMM estimation for the missing data logistic regression model
initial = c(0.2,0.1,0.1,0.1,0.1)
phi.info= est.pi.func( x, "z", NULL, "discrete",dd$u, dd$V, x[,"Y"],
                    x[,"delta"],5 ,
                    initial,method= "L-BFGS-B"
                    
)
```

To analyze the impact of covariates on the time-dependent AUC under nonignorable missing, we apply the following regression model:

\[
     logit{AUC(t;z,\theta)} =  \alpha_{0} t^{0.5} + \alpha_{1} t + \alpha_{0} t^{2} + \beta z
\]

The goal is to estimate the regression coefficients \(\beta\) and make inference on these regression coefficients.

The first step is to reshape the data using the function `data_crossingdc`. The following code reshapes the data and stores the result in the data frame `dd_crossed`. We use a c0 of 1.5 here as an example.

```{r, eval = F}

Y0<-x[,"Y"]   
C0<-x[,"delta"]
M0<-x[,"M"]
VXS<- as.matrix(x[,c("z")]) # ensure this is a matrix
colnames(VXS) <- "z" # ensure to add a name to the column
V0 <- dd$V
datijp <- data_crossingdc(V0,Y0 ,C0,M0,VXS,phi.info$pi.hat,"z",NULL,1.5)


```

The next step is to estimate the regression coefficients using the `fastglm` package. The estimated coefficients are stored in the vector `beta.hat`.

```{r, eval = F}
# install if not installed
# install.packages("fastglm")
library(fastglm)
YI = datijp$yi
XS<-cbind(int = 1,
          #t1 = YI^(-2),
          #t2 = YI^(-1),
          #t3 = YI^(-0.5),
          #t4 = log(YI),
          t5 = YI^(0.5),
          t6 = YI,
          t7 = YI^(2),
          zi = datijp$zi
)


YS<-datijp$Iil
rho01<-datijp$rho01
rho02<-datijp$rho02
rhoweight<- datijp$rhoweight/(datijp$pii * datijp$pil)

ordic = datijp$idi - 1
ordjc = datijp$idl - 1

gc() 


m <- fastglm(XS,YS,weights = rhoweight,family=binomial(link="logit"),maxit=10000L)
beta.hat = m$coefficients
```

The next step is to calculate the asymptotic variance of the estimated regression coefficients. Due to the estimated
weights and the dependence between the two conditional events across all risk sets, the standard asymptotic
theory for maximum likelihood estimation cannot be directly applied to derive the asymptotic properties.
However, if the weight function accounting for missingness is known, the inverse-weighting adjusted estimating
equations in Equation (8) in the manuscript are closely related to a U-statistic of degree 2. Then, the asymptotic variance can be obtained through U-statistic theory and Taylor expansion. The following code calculates the asymptotic variance and stores the result in the variable `V`.

```{r, eval = F}
# extract data used for estimating pi
XI = data.matrix(cbind(int = 1, datijp[,c("yi", "event_indicator_i","zi")], mi = datijp$mi)) 
XL = data.matrix(cbind(int = 1, datijp[,c("yl", "event_indicator_l","zl")], ml = datijp$ml) )
piI = datijp$pii
piL = datijp$pil
L <- covariance_cal_wrapper(beta.hat, M0, rho01, rho02, t(XS), ordic, ordjc, t(XI), t(XL), piI, piL, phi.info$phi.hat, phi.info$composit, phi.info$h.hat)
V <- L$V/n  
```

The Wald statistic and p-value, using the large sample SE, are obtained using the following code:
```{r, eval = F}
SE = sqrt(diag(V))
z_value <- m$coefficients / SE
p_value <- 2 * (1 - pnorm(abs(z_value)))
print(cbind(beta.hat, SE, z_value, p_value))
```

You can also estimate the time-dependent AUC with 95% confidence interval. The following code estimates the time-dependent AUC given z = mean(z) at time t0 ranges from 0.1 to 1:

```{r, eval = F}
t0 <- seq(0.1, 1, by = 0.1)
coeffs = mean(dd$z)
ci <- auc_pred(beta.hat,V,t0,coeffs,  nf=3)
ci
```


